---
title: "in class_assign3"
name: Ashton Cooksey
output:
  html_document: default
  pdf_document: default
---

SUCCESS OF LEADER ASSASSINATION AS A NATURAL EXPERIMENT

One longstanding debate in the study of international relations concerns the
question of whether individual political leaders can make a difference. Some emphasize
that leaders with different ideologies and personalities can significantly affect the
course of a nation. Others argue that political leaders are severely constrained by
historical and institutional forces. Did individuals like Hitler, Mao, Roosevelt, and
Churchill make a big difference? The difficulty of empirically testing these arguments
stems from the fact that the change of leadership is not random and there are many
confounding factors to be adjusted for. 
In this exercise, we consider a natural experiment in which the success or failure
of assassination attempts is assumed to be essentially random.7 Each observation of
the CSV data set leaders.csv contains information about an assassination attempt.
Table 2.8 presents the names and descriptions of variables in this leader assassination data set. The polity variable represents the so-called polity score from the Polity Project. The Polity Project systematically documents and quantifies the regime types of all countries in the world from 1800. The polity score is a 21-point scale ranging from −10 (hereditary monarchy) to 10 (consolidated democracy). The result variable is a 10-category factor variable describing the result of each assassination attempt.

1. How many assassination attempts are recorded in the data? How many countries
experience at least one leader assassination attempt? (The unique() function, What is the average number of such attempts (per year) among these countries?

```{r}
leaders <- read.csv("leaders.csv")
dim(leaders)
summary(leaders$result)
("there were 250 assissination attempts")

unique(leaders$country)
("there are 88 countries who experienced assissination attempts")
mean(tapply(leaders$country, leaders$year, length))
("there are 2.45 attempts per year ")
```

2. Create a new binary variable named success that is equal to 1 if a leader dies
from the attack and 0 if the leader survives. Store this new variable as part of the
original data frame. What is the overall success rate of leader assassination? Does
the result speak to the validity of the assumption that the success of assassination
attempts is randomly determined?

```{r}
result <- factor(leaders$result)
levels(result)
("levels 1-4 are dead. 5-10 are not dead.")

leaders$success[leaders$result == "dies between a day and a week" 
               | leaders$result == "dies between a week and a month" 
               | leaders$result == "dies within a day after the attack"
               | leaders$result == "dies, timing unknown"
               ] <- 1


leaders$success[leaders$result == "hospitalization but no permanent disability" 
               | leaders$result == "not wounded"
               | leaders$result == "plot stopped"
               | leaders$result == "survives but wounded severely"
               | leaders$result == "survives, whether wounded unknown"
               | leaders$result == "wounded lightly"
               ] <- 0
mean(leaders$success)
("there is a 21.6% success rate")
("the success rate seems to not be random. Meaning that it is not valid assumption")
```

3. Investigate whether the average polity score over three years prior to an assassination
attempt differs on average between successful and failed attempts. Also,
examine whether there is any difference in the age of targeted leaders between
successful and failed attempts. Briefly interpret the results in light of the validity
of the aforementioned assumption.

```{r}
tapply(leaders$politybefore, leaders$success, mean)
("The difference is 1.04 which does not appear trivial")
tapply(leaders$age, leaders$success, mean)
("successful aimed leaders is 3 years higher")
```

4. Repeat the same analysis as in the previous question, but this time using the
country’s experience of civil and international war. Create a new binary variable
in the data frame called warbefore. Code the variable such that it is equal to 1
if a country is in either civil or international war during the three years prior to
an assassination attempt. Provide a brief interpretation of the result.

```{r}
leaders$warbefore[leaders$civilwarbefore == 1
                       | leaders$interwarbefore == 1] <- 1 

leaders$warbefore[leaders$civilwarbefore != 1 & leaders$interwarbefore != 1] <- 0 

summary(leaders$warbefore)

war3 <- subset(leaders, leaders$warbefore == 1)

tapply(war3$politybefore, war3$success, mean)
tapply(war3$age, war3$success, mean)
("countries that experience war have a higher polity")
("age of ploted leaders is higher among successes")
```

5. Does successful leader assassination cause democratization? Does successful
leader assassination lead countries to war? When analyzing these data, be sure
to state your assumptions and provide a brief interpretation of the results.

```{r}
leaders$warafter[leaders$civilwarafter == 1
                 | leaders$interwarafter == 1] <- 1

leaders$warafter[leaders$civilwarafter != 1 & leaders$interwarafter != 1] <- 0 

war3b <- subset(leaders, subset = (warafter == 1))

tapply(leaders$warafter, leaders$success, mean)
tapply(leaders$polityafter, leaders$success, mean)
("After assassin success war breaking out is 20%")
("While its rate after failure is 30%")
```

