---
title: "InClassAssign2"
name: Ashton Cooksey
date: "`r Sys.Date()`"
output: html_document
---
Surveys are frequently used to measure political behavior such as voter turnout,
but some researchers are concerned about the accuracy of self-reports. In particular,
they worry about possible social desirability bias where, in postelection surveys,
respondents who did not vote in an election lie about not having voted because they
may feel that they should have voted. Is such a bias present in the American National
Election Studies (ANES)? ANES is a nationwide survey that has been conducted for
every election since 1948. ANES is based on face-to-face interviews with a nationally
representative sample of adults. Table 1.3 displays the names and descriptions of
variables in the turnout.csv data file.

1. Load the data into R and check the dimensions of the data. Also, obtain a
summary of the data. How many observations are there? What is the range of
years covered in this data set?
```{r} 
turnout <- read.csv("turnout.csv")
dim(turnout)
summary(turnout)
("there are 9 columns. Covering years from 1980 to 2008")
```

2. Calculate the turnout rate based on the voting age population or VAP. Note that
for this data set, we must add the total number of eligible overseas voters since
the VAP variable does not include these individuals in the count. Next, calculate
the turnout rate using the voting eligible population or VEP. What difference do
you observe?
```{r}
voter_oversea <- turnout$total / (turnout$VAP +  turnout$overseas) * 100 
(voter_oversea)
names(voter_oversea) <- turnout$year 
voter_oversea
("The numbers represent percent of voting age population that voted")
evoter <- turnout$total / turnout$VEP * 100
names(evoter) <- turnout$year
diff <- evoter - voter_oversea
names(diff) <- turnout$year 
diff
("Presidental elections have higher turnout")
```

3. Compute the differences between the VAP and ANES estimates of turnout rate.
How big is the difference on average? What is the range of the differences?
Conduct the same comparison for the VEP and ANES estimates of voter turnout.
Briefly comment on the results.
```{r}
diffv_o <- turnout$ANES - voter_oversea 
summary(diffv_o)
("on average it is 20.3%, range is 15.1%")
diffev <- turnout$ANES - evoter
summary(diffev)
("for eligible voters the average diff is 16.8%, range is 13.9%")
("there is a large discrepancy between the estimated turnout and actual turnout")
```

4. Compare the VEP turnout rate with the ANES turnout rate separately for
presidential elections and midterm elections. Note that the data set exclude 
5. Divide the data into half by election years such that you subset the data into two
periods. Calculate the difference between the VEP turnout rate and the ANES
turnout rate separately for each year within each period. Has the bias of ANES
increased over time?
```{r}
turnout$year
n.obs <- nrow(turnout)
pres <- c(seq(from =1, to = n.obs, by=2), n.obs)
pevoter <- evoter[pres]
names(pevoter) <- turnout$year[pres]
pevoter
diffpevoter <- turnout$ANES - pevoter
summary(diffpevoter)
("For pres elections, there is a 10.5% difference between the estimated and actual turnout rate on average. This is the closest to ANES")
```
```{r}
mid <- seq(from =2, to = (n.obs -1), by =2)
mevoter <- evoter[mid]
names(mevoter) <- turnout$year[mid]
mevoter
diffmevoter <- turnout$ANES - mevoter
summary(mevoter)
diffmevoter
("There is a huge difference between estimated and actual voter turnout. Meaning very few people vote during midterm elections")
```
# Calculate the difference between the VEP turnout rate and the ANES turnout rate separately for each year within each period. Has the bias of ANES increased over time?

```{r}
old <- turnout[c(1:7), ]
new <- turnout[c(8:14), ]
evotero <- (old$total / old$VEP) * 100 
evotern <- (new$total / new$VEP) * 100 
ANESo <- old$ANES
ANESn <- new$ANES
mean(ANESo) - mean(evotero)
mean(ANESn) - mean(evotern)
("This shows there was an increase in bias for both")
```

6. ANES does not interview prisoners and overseas voters. Calculate an adjustment
to the 2008 VAP turnout rate. Begin by subtracting the total number of ineligible
felons and noncitizens from the VAP to calculate an adjusted VAP. Next,
calculate an adjusted VAP turnout rate, taking care to subtract the number of
overseas ballots counted from the total ballots in 2008. Compare the adjusted
VAP turnout with the unadjusted VAP, VEP, and the ANES turnout rate. Briefly
discuss the results.
```{r}
VAPc <- (turnout$VAP - turnout$felons - turnout$noncit)
VAPc
turnout08 <- turnout[14, ]
turnouttotal <- turnout08$total - turnout08$osvoters
turnouttotal
adjVAP <- turnout$total / VAPc
adjVAP
("break")
mean(adjVAP) * 100

```

